---
title: "Reproducible Research - Project 1"
author: "Kier O'Neil"
date: "November 29, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Executive Summary

In this assignment I was asked to analyze a set of activity data taken from a single subject over the span of Oct 1 to Nov 30, 2012.   

# Requirements:
1. Code for reading in the dataset and/or processing the data
2. Histogram of the total number of steps taken each day
3. Mean and median number of steps taken each day
4. Time series plot of the average number of steps taken
5. The 5-minute interval that, on average, contains the maximum number of steps
6. Code to describe and show a strategy for imputing missing data
7. Histogram of the total number of steps taken each day after missing values are imputed
8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
9. All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

# Preliminary work
## Load libraries  

```{r message=FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(lubridate)
```

## Load Data
```{r}
activity_raw <- read_csv("activity.csv")
```

## Exploration

Here are the first six rows.  

```{r}
head(activity_raw)
```

We can see that there are at least a few missing values

###How many total missing values?

```{r}
colSums(is.na(activity_raw))
```

There are 2304 missing values, all in steps variable, of 17568 records.  This is 13% of the entire dataset.  


### What is that interval variable used for?
```{r}
range(activity_raw$interval)
```

It looks like the interval is part of the date.  I will assume that it is a 5-minute increment of the date and should be in 00:00 format.

### What is the date range?
```{r}
range(activity_raw$date)
```
Only about two months worth of data. 

### Here is how the an average day looked over the entire two months
```{r}
act_plot <- na.omit(activity_raw) %>%
    group_by(interval) %>%
    summarise(avg_steps = mean(steps))

ggplot(act_plot, aes(x=interval, y=avg_steps)) + geom_line()
    

```

This person gets up around 5:15am has a big spike in activity around 8:30am, has a fairly active but constant activity pattern from 10am to about 7:30pm and then a drastic reduction in average activity until the next morning.  


# Preprocessing  

Process/transform the data (if necessary) into a format suitable for your analysis.  

### Which days have NA's
```{r}
activity_raw %>%
    group_by(date) %>%
    filter(is.na(steps)) %>%
    summarise(num_of_na = length(steps))

```

Interesting.  The first and last days contain mostly NA's. Let's trim them off.  
```{r}
ar_trimmed <- activity_raw %>%
    mutate(date = ymd(date)) %>%
    filter(date >= "2012-10-01", date < "2012-11-30")
```
# Question 1  
## What is mean total number of steps taken per day?

For this part of the assignment, you can __ignore the missing values in the dataset__.

1. Calculate the total number of steps taken per day
2. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

### Step 1: Ignore missing values ;-)
```{r}
ar_na_omit <- na.omit(activity_raw)
# Boom!!
```

### Step 2:  Calculate the total number of steps taken per day.

This is where I like to use dplyr.  
```{r}
ar_grouped_summed <- ar_na_omit %>%
    group_by(date) %>%
    summarise(total_steps = sum(steps))
``` 

### Step 3: Make a histogram of total number of steps taken per day.  

This is where I like to use ggplot.  
```{r}
ggplot(ar_grouped_summed, aes(x= total_steps)) + geom_histogram(binwidth = 1250)
```

### Step 4: Calculate and report the mean and median. 

```{r}
ar_mean_median <- ar_grouped_summed %>%
    summarise(mean_steps = mean(total_steps), median_steps = median(total_steps))
ar_mean_median
# Booyah
```
Pretty darn close.  

# Question 2  

## What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis). 
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

### Step 1:  Get data we need

Let's switch the dataset around to be grouped by the average of the 5 minute interval.
```{r}
ar_5m <- ar_na_omit %>%
    group_by(interval) %>%
    summarise(mean_steps = mean(steps))
```

### Step 2: Plot interval on x and average steps on y  
```{r}
ggplot(ar_5m, aes(x = interval, y = mean_steps)) + geom_line()

```

### Step 3: Which 5-minute interval had the highest average.  
```{r}
max_interval <- which(ar_5m$mean_steps == max(ar_5m$mean_steps))

#show the interval with the highest average steps 
ar_5m[max_interval, ]
```

Our subject has the most average steps in the time interval of 8:35am, with 206.17 steps on average in those five minutes each day.

This person sleeps until about 5:15 and gradually ramps up to most activity around 8:35am.
During the day this subject maintains around 25 to 100 steps per hour then slows down after 6pm.  

# Question 3  
## Imputing missing values

*Note that there are a number of days/intervals where there are missing values (coded as NAs). The presence of missing days may introduce bias into some calculations or summaries of the data.*

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

### Step 1: Let's switch back to the full data set but trim off the first and last day.  
Even though they contain NAs that could be imputed they are at the edges and serve no useful purpose.

We're also going to add in a weekday variable to break up the days of the week.  This is foreshadowing for later.  
```{r}
ar1 <- activity_raw %>%
    filter(date != "2012-10-01" & date != "2012-11-30") %>%
    mutate(weekday = factor(wday(date)))
```

### Step 2: Let's figure out how many NAs are in this dataset
The question specifically asks how many rows have missing values.  
```{r}
summary(ar1)
```

We can see that there are 1728 rows with NA values in steps.

That is 10% of the values.

## Let's get ready to model and apply
### Step 1: Break data into train and test sets

Training set has step values, and Test set has NAs.  

```{r}
ar_train <- ar1[!is.na(ar1$steps), ]
ar_test <- ar1[is.na(ar1$steps), ]
```

## Strategy for imputing missing values

Use the linear model function ("lm") to build a model where steps will be derived from which day of the week and which time of the day were actually observed.  

### Step 2: Generate a model to predict steps based on weekday and interval
```{r}
fit <- lm(data = ar_train, steps ~ factor(weekday) + interval)
```

###Step 3: Apply prediction model to test set to impute steps.
```{r}
ar_test$steps <- predict(fit, ar_test)
```

###Step 4: Now bind it all back together  
```{r}
ar_imputed <- rbind(ar_train, ar_test)
```

###Step 5: Do sanity check that new dataset has same number of rows as old dataset.  
```{r}
dim(ar1)
dim(ar_imputed)
summary(ar_imputed)
```
All NA's have been replaced by the predictions from the fit model.

Cool, eh!!  (fulfilling requirement #3 ;-)

# Question 4

##Are there differences in activity patterns between weekdays and weekends?

For this part the I use the weekdays() function with the imputed dataset.
### Step 1: Create a weekday variable

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
# Label weekdays
ar_full_weekday <- ar_imputed %>%
    filter(!weekday %in% c(1,7)) %>%
    mutate(day_type = "weekday")
ar_full_weekend <- ar_imputed %>%
    filter(weekday %in% c(1,7)) %>%
    mutate(day_type = "weekend")
```

### Step 2: Join the two set together
```{r}
ar_w_daytype <- rbind(ar_full_weekday, ar_full_weekend)
ar_w_daytype$day_type <- factor(ar_w_daytype$day_type, 
                                levels = c("weekday", "weekend"))
```

### Step 3: Summarise average steps per interval based on day_type

```{r}
total_steps_days <- ar_w_daytype %>%
    group_by(day_type, interval) %>%
    summarise(avg_steps = mean(steps))
```

### Step 4: Make a panel plot with two plots showing different activity patterns for weekday vs weekend.  

It is a time series line-plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
p <- ggplot(total_steps_days, aes(interval, avg_steps)) + 
    geom_line() + geom_smooth()
p + facet_wrap(~day_type, ncol=2, labeller = "label_both")

```